---
title: "Appendix"
author: "Max Aantjes"
date: "22/08/2020"
output:
  github_document:
    toc: true
    toc_depth: 3
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Packages
The following R packages are used in this analysis.

```{r, message = FALSE}
library(tidyverse)
library(knitr)
library(stringr)
library(ggplot2)
```
# Data Extraction
## Data Source and Relevance
The indicator values for this analysis are taken from the results of the  Ecuadorian National Multi-Purpose Household Survey of December 2019 (Encuesta Nacional Multipropósito de Hogares - Diciembre 2019). The survey collected data for a variety of socio-economic indicators and was conducted by the Ecuadorian national census and statistics institute, INEC. At the time of writing, the form and raw data are available in this [online repository](https://www.ecuadorencifras.gob.ec/encuesta-nacional-multiproposito-de-hogares/). 

## Download
The following code is used to extract the data from the INEC's official website. 

```{r echo = TRUE, results = 'hide', message = FALSE}
url <- "https://www.ecuadorencifras.gob.ec/documentos/web-inec/Multiproposito/2019/BDD_DATOS_ABIERTOS_MULTI%20_2019_12_CSV.zip"
download.file(url, "temp.zip")
unzip("temp.zip")
dfhousehold <- read.csv("BDD_DATOS_ABIERTOS_MULTI _2019_12_CSV//201912_multibdd_equipamiento.sav.csv", sep = ";")
dfperson <-  read.csv("BDD_DATOS_ABIERTOS_MULTI _2019_12_CSV//201912_multibdd_personas.sav.csv", sep = ";")
unlink("BDD_DATOS_ABIERTOS_MULTI _2019_12_CSV", recursive = TRUE)
file.remove("temp.zip")
```

# Data Cleaning
## Missing Values
Missing values will not be removed at this stage for the ```dfperson``` and ```dfequipment``` data sets. The reason for this is that the questions in their respective survey sections were constructed in a hierarchical manner, where answers to earlier questions can make later questions redundant. Follow-up questions (such as for example: do you own a smartphone?) were not asked if the respondent answered "no"to a preceding question (such as for example: do you own a phone?). For this reason, missing values do **not** necessarily indicate missing information. Accordingly, missing values are interpreted, adapted and potentially removed at sub-indicator level, on a case by case basis. The code and rational for this process are available in the other appendix sections.  

## Merging the Two Data Sets
The information on household level and person level is first merged using the *id_hogar* variable. 

```{r}
df <- dfperson %>%
        left_join(dfhousehold, by = "id_hogar") %>%
        rename("area" = 1, "postcode" = 2)
```

## Subset of Data of Interest
Only households with children or young people of schoolgoing age (5 - 18 year-olds) will be considered in this analysis. For this purpose, the data is subsetted. The ```agecheck()``` function checks if there is at least one household member who is between 5 and 18 years old, returning value 1, or not, returning value 0. 

```{r}
agecheck <- function(x){
        y <- 0
        for(i in 1:length(x)){
                if(x[i] %in% 5:18){
                        y <-1}}
        return(y)}
```

The households identification numbers of household where the ```agecheck()``` function returns 1 are then stored in the ```subset``` data frame.  
```{r}
subset <- df %>%
        group_by(id_hogar) %>%
        summarise(subset = agecheck(s1p3)) %>%
        select(id_hogar, subset)
```

The households of interest are subsetted from the ```dfequipment``` data frame. 

```{r}
df <- df %>%
        left_join(subset, by = "id_hogar") %>%
        filter(subset == 1) %>%
        select(-subset)
```


# Analysis
## Availability of Electronic Devices
This sub-indicator estimates relative risk based on the average number of electronic devices per household member. This indicator can be estimated through 8 different variables:

```{r echo = FALSE, results = 'asis'}
names <- c("Survey", "Code", "Form Section", "Question", "Responses")
survey <- rep("National Multi-Purpose Household Survey of December 2018", 9)
code <- c("s1p3", "s7p1a", "s7p1b", "s7p1a1", "s11p11a", "s11p11b", "s11p12a", "s11p12b", "id_hogar")
form.section <- c("Section 1, question 3", "section 7, question 1, subsection 1", "section 7, question 1, subsection 2", "section 7, question 1A", "Section 11, Question 1, subsection 1, subsection 'code'", "Section 11, Question 1, subsection 1, subsection 'number'", "Section 11, question 1, subsection 2, subsection 'code'", "section 11, question 1, subsection 2, subsection 'number'", "-")
question <- c("How old are you? ¿Cuántos años cumplidos tiene (…)?", "Do you have an activated phone? (¿(…) tiene teléfono celular activado?)", "How many activated phones do you have? (¿Cuántos?)", "Is your phone a smartphone? (¿El (los) teléfono (s) celular (es) que tiene (…..) es/son smartphone (teléfono Inteligente, se puede comunicar a través de Wi-Fi, e-mails, red de datos, etc.)?)", "Does the household have a desktop computer? (¿Tiene este hogar: computadora de escritorio?)", "How many desktop computers does the household have? (¿Cuántos tiene?)", "Does the household have a laptop or tablet? (¿Tiene este hogar: computadora laptop/ tablet?)", "How many laptops and tablets does the household have? (¿Cuántos tiene?)", "Identification code for household assigned by INEC")
responses <- c("numeric", "1 = yes, 2 = no", "numeric", "1 = yes, 2 = no", "1 = yes, 2 = no", "numeric", "1 = yes, 2 = no", "numeric", "string of characters (numbers)")
table <- data.frame(cbind(survey, code, form.section, question, responses))
names(table) <- names
kable(table, caption = "Table 2.1: Relevant Variables for Indicator 2.1")
```

## Subset of Data of Interest
1-4 year olds are removed from the data set as they do not need to use electronic devices nor are expected to use electronic devices.

```{r}
df1 <- df %>%
        filter(s1p3 > 4) %>%
        select(s1p3, s7p1a, s7p1b, s7p1a1, s11p11a, 
               s11p11b, s11p12a, s11p12b, id_hogar)
```

## Number of smartphones per household
The code transforms the values of variable *s7p1a* as follows: value 2 (which means the respondent does not have an activated phone) is replaced with 0; value 1 (which means the respondent does have an activated phone) stays 1. There were no NA values.
```{r}
df1$s7p1a[df1$s7p1a == 2] <- 0
sum(is.na(df1$s7p1a))
```

The code then assigns the value 0 to variable *s7p1b* (the number of phones the respondent has) if *s7p1a* equals 2 (which means the respondent does not own any activated phone). In other cases the value of *s7p1b* is not changed. There are no NA values.

```{r}
df1$s7p1b[df1$s7p1a == 0] <- 0
sum(is.na(df1$s7p1b))
```

Next, the code replaces the following values of variable *s7p1a1*: value 2 (which means the respondent's phone(s) is/are **not** smartphone(s)) is replaced with 0; value 1 (which means the respondent's phone(s) is/are smartphone(s)) remains 1. For rows where *s7p1b* equals 0 (which means the respondent does not own any phones), the value of *s7p1a1* is also set to 0 (which means the respondent does not own any smartphones). There are no NA values.

```{r}
df1$s7p1a1 <- ifelse(df1$s7p1a1 == 1, 1, 0)
df1$s7p1a1[df1$s7p1b == 0] <- 0
sum(is.na(df1$s7p1a1))
```
Following these operations, the number of smartphones per individual can now be calculated by multiplying *s7p1b* (which indicates the number of activated phones) with *s7p1a1* (which indicates if the respondent has/ does not have a smartphone). Finally, the number of smartphones is summed per household using the *id_hogar* variable. The results are stored in the ```dfphone``` data set.  

```{r}
dfphone <- df1 %>%
        mutate(phone = s7p1b * s7p1a1) %>%
        group_by(id_hogar) %>%
        summarise(no.smartphones = sum(phone))
```

## Access To Internet

The access to internet can be estimated with the following variables:

```{r echo = FALSE}
names <- c("Variable", "Form Section", "Question", "Responses")
variable <- c("area", "s11p2")
form.section <- c("NA", "Section 11, question 2")
question <- c("Determined by Interviewer", "Does the household have access to internet? (Tiene este hogar acceso a internet?)")
responses <- c("1 = rural, 2 = urban", "1 = yes, 2 = no")
table <- data.frame(cbind(variable, form.section, question, responses))
names(table) <- names
kable(table)
```

The variables are first converted into factor variables. There are no NA values in the data frame.

```{r}
df1 <- df %>%
        mutate(s11p2 = factor(s11p2, levels = 1:2, labels = c("Have Access", "Have No Access"))) %>%
        mutate(area = factor(area, levels = 1:2, labels = c("urban", "rural")))
sum(is.na(df$s11p2))
```

The data from respondents aged 5 to 18 is then filtered.

```{r}
df2 <- df1 %>%
        filter(s1p3 %in% 5:18)
nrow(df2)
```

Subsequently, the proportions of respondents with and without access to the internet are calculated for both areas (rural and urban). This is saved in a new data frame.

```{r}
intacc <- df2 %>%
        group_by(area, s11p2) %>%
        count() %>%
        ungroup() %>%
        group_by(area) %>%
        mutate(prop = n/sum(n)*100) %>%
        select(area, s11p2, prop)
```

Finally, a graph is created which compares the proportions in rural and urban areas. 

```{r}
p <- ggplot(data = intacc, aes(y = prop, x = area, fill = s11p2)) + theme_bw()
p1 <- p + geom_bar(position = "dodge", stat = "identity") 
p2 <- p1 + labs(y = "Proportion of respondents (in %)", x = "", title = "Figure 2: Lack of Home Internet Connection of School-Age Population", subtitle = "Based on data from 10,543 respondents between 5 and 18 years old. Surveyed\nhouseholds were selected through randomised cluster sampling at national level,\nwhere clusters represented different socio-economic strata.", caption = "Source: INEC Multi-Purpose Household Survey 2019", fill = "")
p3 <- p2 + scale_fill_manual(name = "", values = (c("Have Access" = "#A3BFA5", "Have No Access" = "#A26E6E")))
p3
```

Figure 1.1 graphs the resulting distribution. The distribution is is slightly left-skewed. Nevertheless, the distance between the minimum and maximum values remain meaningful in terms of relative risk: the minimum value, 27%, indicates a substantial amount of respondents in the province were connected (which corresponds with the lowest level of risk); the maximum value, 81%, indicates almost none of the survey respondents in the province were connected (which corresponds with the highest level of risk). There are thus no problematic outliers and normalisation can be performed.  

```{r echo = FALSE}
median <- median(dim1ind1$access)
p <- ggplot(data = dim1ind1, aes(x = access)) + theme_bw()
p1 <- p + geom_histogram(binwidth = 5)
p2 <- p1 + geom_vline(aes(xintercept = median, col = "Median"))
p3 <- p2 + labs(x = "Proportion of respondents not connected\nat home per Province (in %)", y = "Frequency", title = "Figure 1.1: Lack of Home Internet Connection per Province", color = "")
p3
```

## Digital Literacy


