---
title: "Appendix"
author: "Max Aantjes"
date: "22/08/2020"
output:
  github_document:
    toc: true
    toc_depth: 2
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
The following sections contain all computations for the creation of the exploratory data results summarised in the brief paper. Computations are accompanied by references and justifications to clarify any choices made. Through *code book tables*, explicit references are also made to the survey questions which responses were used in the analysis. 

# R Packages
The following R packages are used in the subsequent code sections. 

```{r, message = FALSE}
library(tidyverse)
library(knitr)
library(ggplot2)
```
# Data Extraction
### Data Source and Relevance
The analysis performed below is based on the results of the Ecuadorian National Multi-Purpose Household Survey of December 2019 (Encuesta Nacional Multipropósito de Hogares - Diciembre 2019). The survey collected data for a variety of socio-economic indicators and was conducted by the Ecuadorian national census and statistics institute, INEC. At the time of writing, the form and raw data are available in this [online repository](https://www.ecuadorencifras.gob.ec/encuesta-nacional-multiproposito-de-hogares/). According to INEC, the data is representative at national, national-rural and national-urban level. A total of 11,528 households were surveyed, with the sampling design taking the form of randomised cluster sampling. 

Supplementary information on the data collection process and the definition of some variables has been made available by INEC (Andrade et al., 2020). An example of an important variable defined in that document is the distinction between rural areas, i.e. population centres of less than 2,000 citizens, viz. urban areas, i.e. population centres of 2000 citizens or more (ibid, p. 11). All other interpretations of variables are directly taken from the form used during the survey process (INEC, 2019).

### Download
The following code is used to extract the data from the data repository. 

```{r echo = TRUE, results = 'hide', message = FALSE}
url <- "https://www.ecuadorencifras.gob.ec/documentos/web-inec/Multiproposito/2019/BDD_DATOS_ABIERTOS_MULTI%20_2019_12_CSV.zip"
download.file(url, "temp.zip")
unzip("temp.zip")
dfhousehold <- read.csv("BDD_DATOS_ABIERTOS_MULTI _2019_12_CSV//201912_multibdd_equipamiento.sav.csv", sep = ";")
dfperson <-  read.csv("BDD_DATOS_ABIERTOS_MULTI _2019_12_CSV//201912_multibdd_personas.sav.csv", sep = ";")
unlink("BDD_DATOS_ABIERTOS_MULTI _2019_12_CSV", recursive = TRUE)
file.remove("temp.zip")
```

# Data Cleaning
### Removal of Missing Values
Missing values are not removed at this stage. The questions in the survey were constructed in a hierarchical manner, where answers to earlier questions can make later questions redundant. Follow-up questions (such as for example: do you own a smartphone?) were not asked if the respondent answered "no" to a preceding question (such as for example: do you own a phone?). For this reason, missing values do **not** necessarily indicate missing information. Accordingly, missing values are interpreted, adapted and potentially removed on a case by case basis.

### Specification of Variable Classes
First the classes of the categorical variables which are used for grouping and identification are specified. The relevant variables are first summarised in *Code Book Table 1*:

```{r echo = FALSE, results = 'asis'}
names <- c("Code", "Form Section", "Question", "Responses")
code <- c("id_per", "id_hogar", "area", "s1p9", "s1p3")
form.section <- c("-", "-", "-", "Section 1, question 9", "Section 1, question 3")
question <- c("Identification code for person assigned by INEC", "Identification code for household assigned by INEC", "Determined by Interviewer", "How do you identify yourself according to your culture and customs (¿Cómo se identifica (…) según su cultura y costumbres:?)", "How old are you? ¿Cuántos años cumplidos tiene (…)?")
responses <- c(rep("string of characters (numbers)", 2), "1 = urban, 2 = rural", "1 =indigenous, 2 = afroecuadorian, 3 = black, 4 = mulato, 5 = montuvio, 6 = mestizo, 7 = white, 8 = other", "0 = less than 1 year old, 1 = 1 year old, etc.")
table <- data.frame(cbind(code, form.section, question, responses))
names(table) <- names
kable(table, caption = "Code Book Table 1")
```

The subsequent code transforms each of these variables into factor variables in line with *Code Book Table 1*. 

```{r}
dfperson <- dfperson %>%
        rename("area" = 1) %>%
        mutate(id_per = as.character(id_per)) %>%
        mutate(id_hogar = as.character(id_hogar)) %>%
        mutate(area = factor(area, levels = 1:2, labels = c("urban", "rural"))) %>%
        mutate(eth = factor(s1p9, levels = 1:8, labels = c("indigenous", "afroecuadorian", "black", "mulato", 
                                                           "montuvio", "mestizo", "white", "other")))
        
dfhousehold <- dfhousehold %>%
        mutate(id_hogar = as.character(id_hogar))
```

There were no missing values recorded in terms of these four variables, apart from the ```eth``` variable.

```{r}
sapply(list(is.na(dfhousehold$id_hogar), is.na(dfperson$id_per), is.na(dfperson$id_hogar), 
            is.na(dfperson$area), is.na(dfperson$eth)), sum)
```

The 2,690 missing values in the ```eth``` variable were all recorded with 0-4 year-olds. Since these respondents are not of school-going age nor likely compete for resources such as electronic devices needed by the school-going population, they are not of relevance to the data analysis. Accordingly, this age group will later be removed from the data set. 

```{r}
unique(dfperson$s1p3[is.na(dfperson$eth)])
```

### Merge of the Two Data Sets
The information on household level and person level can now be merged using the *id_hogar* variable. 

```{r}
df <- dfperson %>%
        left_join(dfhousehold, by = "id_hogar") 
```

### Subset of the Data of Interest
Only households with children or young people of schoolgoing age (5 - 18 year-olds) are relevant to this analysis. For this purpose, a subset is taken from the data. The ```agecheck()``` function checks if there is at least one household member who is between 5 and 18 years old, returning value 1, or not, returning value 0. 

```{r}
agecheck <- function(x){
        y <- 0
        for(i in 1:length(x)){
                if(x[i] %in% 5:18){
                        y <-1}}
        return(y)}
```

The households identification numbers of household where the ```agecheck()``` function returns 1 are then stored in the ```subset``` data frame.  
```{r}
subset <- df %>%
        group_by(id_hogar) %>%
        summarise(subset = agecheck(s1p3)) %>%
        select(id_hogar, subset)
```

A subset of the households of interest is taken from the ```df``` data frame. For reasons explained above, any respondents between 0 and 4 years old are also removed from the data set. 

```{r}
df <- df %>%
        left_join(subset, by = "id_hogar") %>%
        filter(subset == 1) %>%
        filter(s1p3 > 4) %>%
        select(-subset)
```

# Analysis
### Availability of Smart Devices
#### Code Book
The first graph aims to estimate the relative risk of exclusion from education based on the average number of electronic devices per household member. This graph relies on 7 different variables summarised in *Code Book Table 2*: 
```{r echo = FALSE, results = 'asis'}
names <- c("Code", "Form Section", "Question", "Responses")
code <- c("s7p1a", "s7p1b", "s7p1a1", "s11p11a", "s11p11b", "s11p12a", "s11p12b")
form.section <- c("section 7, question 1, subsection 1", "section 7, question 1, subsection 2", "section 7, question 1A", "Section 11, Question 1, subsection 1, subsection 'code'", "Section 11, Question 1, subsection 1, subsection 'number'", "Section 11, question 1, subsection 2, subsection 'code'", "section 11, question 1, subsection 2, subsection 'number'")
question <- c("Do you have an activated phone? (¿(…) tiene teléfono celular activado?)", "How many activated phones do you have? (¿Cuántos?)", "Is your phone a smartphone? (¿El (los) teléfono (s) celular (es) que tiene (…..) es/son smartphone (teléfono Inteligente, se puede comunicar a través de Wi-Fi, e-mails, red de datos, etc.)?)", "Does the household have a desktop computer? (¿Tiene este hogar: computadora de escritorio?)", "How many desktop computers does the household have? (¿Cuántos tiene?)", "Does the household have a laptop or tablet? (¿Tiene este hogar: computadora laptop/ tablet?)", "How many laptops and tablets does the household have? (¿Cuántos tiene?)")
responses <- c("1 = yes, 2 = no", "numeric", "1 = yes, 2 = no", "1 = yes, 2 = no", "numeric", "1 = yes, 2 = no", "numeric")
table <- data.frame(cbind(code, form.section, question, responses))
names(table) <- names
kable(table, caption = "Code Book Table 2")
```

#### Number of Smartphones per Household
The code first assigns the value 0 to variable *s7p1b* (the number of phones the respondent has) if *s7p1a* equals 2 (which means the respondent does not own any activated phone). In other cases the value of *s7p1b* is not changed. There were no missing values in variable *s7p1a* nor in *s7p1b*.

```{r}
df1 <- df
df1$s7p1b[df1$s7p1a == 2] <- 0
sapply(list(is.na(df1$s7p1b),is.na(df1$s7p1a)), sum)
```

Next, the code replaces the following values of variable *s7p1a1*: value 2 (which means the respondent's phone(s) is/are **not** smartphone(s)) is replaced with 0; value 1 (which means the respondent's phone(s) is/are smartphone(s)) remains 1. For rows where *s7p1b* equals 0 (which means the respondent does not own any activated phones), the value of *s7p1a1* is also set to 0 (which means the respondent does not own any smartphones). There were no missing values in variable *s7p1a1*. 

```{r}
df1$s7p1a1 <- ifelse(df1$s7p1a1 == 1, 1, 0)
df1$s7p1a1[df1$s7p1b == 0] <- 0
sum(is.na(df1$s7p1a1))
```

Following these operations, the number of smartphones per individual can now be calculated by multiplying *s7p1b* (which indicates the number of activated phones) with *s7p1a1* (which indicates if the respondent's phone(s) is/are smartphone(s)). Finally, the number of smartphones is summed per household using the *id_hogar* variable. The results are stored in the ```dfphone``` data set.  

```{r}
dfphone <- df1 %>%
        mutate(phone = s7p1b * s7p1a1) %>%
        group_by(id_hogar) %>%
        summarise(no.smartphones = sum(phone))
```

#### Number of computers, laptops and tablets per household
The following code creates a data frame listing the number of computers, laptops and tablets per household. It first sets the value of variable *s11p11b* (the number of desktop computers) to 0 if *s11p11a* equals 2 (which means the household does not have a desktop computer). There were no missing values in variable *s11p11a* nor in *s11p11b*.

```{r}
df1$s11p11b[df1$s11p11a == 2] <- 0
sapply(list(is.na(df1$s11p11ab),is.na(df1$s11p11b)), sum)
```

Next, the code sets the value of variable *s11p12b* (the number of tablets and laptops) to 0 if *sp11p12a* equals 2 (which means the household does not have a tablet nor laptop). There were no missing values in variable *s11p12a* nor in *s11p12b*.

```{r}
df1$s11p12b[df1$s11p12a == 2] <- 0
sapply(list(is.na(df1$s11p12ab),is.na(df1$s11p12b)), sum)
```

Finally, variables *s11p11b* (the number of desktop computers) and variable *s11p12b* (the number of laptops and tablets) are summed into the *no.computers* variable. The duplicate values for each household (which exist because we have calculated the household value for each household member) are then removed. The results are stored in the ```dfcomputer``` data set. 

```{r}
dfcomputer <- df1 %>%
        mutate(no.computers = s11p11b + s11p12b) %>%
        select(id_hogar, no.computers) %>%
        distinct()
```

#### Number of Devices per Household Member
The following code creates a data set, ```dfmember``` which indicates for each household the number of members of age 5 and older.

```{r}
dfmember <- df1 %>%
        count(id_hogar)
```

The following code creates a data set, ```dfarea``` which indicates the area per household.

```{r}
dfarea <- df1 %>%
        select(id_hogar, area) %>%
        distinct()
```

The following code calculates for each household how many electronic devices are available for each of its 5+ year-old members. For this purpose, the dfcomputer and dfphone data sets are first merged through the *id_hogar* variable. The variables *computer* (which indicates the number of desktop computers, laptops and tablets in the household) and the variable *phone* (which indicates the number of smartphones in the household) are subsequently summed into the variable *no.devices*. 

Next, the code merges the resulting data set with the dfmember data set through the *id_hogar* variable. The number of devices per household member are then calculated by dividing the *no.devices* (which indicates the total number of electronic devices) with  *n* (which indicates the number of household members). The results are stored in the *ratio* variable. The area is then matched with households by merging the data set with the ```dfarea``` data set. 

```{r}
dfdevices <- dfcomputer %>%
        left_join(dfphone, by = "id_hogar") %>%
        mutate(no.devices = no.computers + no.smartphones) %>%
        left_join(dfmember, by = "id_hogar") %>%
        mutate(ratio = no.devices/n) %>%
        left_join(dfarea, by = "id_hogar")
```

We can now calculate the mean, median and standard deviation values of the *ratio* variable for rural and urban areas.

```{r}
dfdevices %>%
  group_by(area) %>%
  summarise(mean(ratio), median(ratio), sd(ratio))
```

We can also calculate the number of households surveyed per area. There were 3,595 urban households and 2,254 rural households.

```{r}
summary(dfarea$area)
```

Finally, a graph is created which compares the distribution of device availability across urban and rural regions. 1 outlier is removed from the graph for readability purposes. 

```{r}
p <- ggplot() + theme_bw()
p1 <- p + geom_boxplot(data = dfdevices, aes(y = ratio, x = area, fill = area)) + ylim(0,3)
p2 <- p1 + geom_hline(aes(yintercept = 1, col = "Ideal\nNumber of\nDevices"), size = 1.2, lty = "dashed") 
p3 <- p2 + labs(y = "Number of Smart Devices per Household\nMember of 5+ years old", x = "", 
                title = "Figure 1: Availability of Smart Devices per Ecuadorian\nHousehold Member", 
                subtitle = "Based on data from 3,595 urban households and 2,254 rural households with\nchildren and young people of school-going age (5-18 year-olds). Devices counted\nas smart devices are: smartphones, laptops, tablets and desktops.", 
                caption = "Source: Data from INEC Multi-Purpose Household Survey 2019\nData Summary by author (Max Aantjes)", fill = "")
p4 <- p3 + scale_color_manual(name = "", values = c("Ideal\nNumber of\nDevices" = "#E96519")) + scale_fill_manual(name = "", values = c("urban" = "#907D73", "rural" = "#C1C37F"))
p4
```

### Access to Internet
#### Code Book
The second graph aims to estimate the relative risk of exclusion from education based on the access to internet in households. This graph relies on 1 variable summarised in *Code Book Table 3*: 

```{r echo = FALSE, results = 'asis'}
names <- c("Variable", "Form Section", "Question", "Responses")
variable <- c("s11p2")
form.section <- c("Section 11, question 2")
question <- c("Does the household have access to internet? (¿Tiene este hogar acceso a internet?)")
responses <- c("1 = yes, 2 = no")
table <- data.frame(cbind(variable, form.section, question, responses))
names(table) <- names
kable(table, caption = "Code Book Table 3")
```

#### Learner Access to Internet at Home
The variable of interest is first converted into a factor variable in line with *Code Book Table 3*. There are no missing values in the data frame.

```{r}
df1 <- df %>%
        mutate(s11p2 = factor(s11p2, levels = 1:2, labels = c("Have Access", "Do not have Access"))) 
sum(is.na(df$s11p2))
```

A subset of the respondents aged 5 to 18 years old is then taken from the data set. The subset contains responses from 1083 indigenous, 165 afroecuadorian, 175 black, 135 mulato, 541 montuvio, 8267 mestizo, 172 white and 5 'other' respondents. Without 'other', the resulting data set has 10,538 respondents. 

```{r}
df2 <- df1 %>%
        filter(s1p3 %in% 5:18)
summary(df2$eth)
nrow(df2) - 5
```

Subsequently, the proportions of respondents with and without access to the internet are calculated for each ethnic group. This is saved in a new data frame, ```dfacc```. The ethnic group, 'other', is removed due to the small sample size (<6). Each ethnic group is then assigned a rank in accordance with the proportion of respondents having access to internet. 

```{r}
dfacc <- df2 %>%
        filter(eth != "other") %>%
        group_by(eth, s11p2) %>%
        count() %>%
        ungroup() %>%
        group_by(eth) %>%
        mutate(prop = n/sum(n)*100)

order <- dfacc %>%
        filter(s11p2 == "Have Access") %>%
        arrange(prop) %>%
        rowid_to_column("rank") %>%
        select(eth, rank)

dfacc <- dfacc %>%
        left_join(order, by = "eth")
```

The results are listed in *table 1*. 

```{r echo = FALSE, results = 'asis'}
table <- dfacc %>%
        mutate(prop = paste0(round(prop), "%")) %>%
        pivot_wider(names_from = s11p2, values_from = c(prop, n)) %>%
        arrange(rank) %>%
        select(-rank)
kable(table, caption = "Table 1: Home Internet Access per Ethnic group")
```

Finally, a stacked bar chart is created which compares the proportions in terms of different ethnic groups. 

```{r}
p <- ggplot(data = dfacc, aes(y = prop, x = reorder(eth, rank), fill = s11p2)) + theme_bw()
p1 <- p + geom_bar(position = "stack", stat = "identity") 
p2 <- p1 + labs(y = "Proportion of respondents (in %)", x = "", 
                title = "Figure 2: Home Internet Connection of Ecuadorian\nSchool-Age Population per Self-identified Ethnicity", 
                subtitle = "Based on data from 10,538 respondents between 5 and 18 years old. Sample\nsizes across ethnic groups varied, with 541 montuvio, 1083 indigenous, 165\nafroecuadorian, 175 black, 135 mulato, 8267 mestizo and 172 white respondents.", 
                caption = "Source: Data from INEC Multi-Purpose Household Survey 2019\nData Summary by author (Max Aantjes)", fill = "") + 
        theme(axis.text.x = element_text(angle = 90, vjust = 0.2))
p3 <- p2 + scale_fill_manual(name = "", values = (c("Have Access" = "#A3BFA5", "Do not have Access" = "#A26E6E")))
p3
```

### Digital Literacy
#### Code Book
The third graph aims to estimate the relative risk of exclusion from education based on the digital literacy of computer users. This graph relies on 7 variables summarised in *Code Book Table 4*: 

```{r echo = FALSE, results = 'asis'}
names <- c("Code", "Form Section", "Question", "Responses")
code <- c("s7p6", "s7p71", "s7p72", "s7p73", "s7p75", "s7p76", "s7p78", "s7p79")
form.section <- c("Section 7, Question 6", "Section 7, question 7, subsection 1", "Section 7, question 7, subsection 2", "Section 7, question 7, subsection 3", "Section 7, question 7, subsection 5", "Section 7, question 7, subsection 6", "Section 7, question 7, subsection 7", "Section 7, question 7, subsection 8")
question <- c("Have you used a computer from anywhere in the past 12 months?", "In the past 12 months when using a computer or laptop: have you copied or moved a document or folder?", "In the past 12 months when using a computer or laptop: have you used the copy and paste function in a document?", "In the past 12 months when using a computer or laptop: have you sent an email with an attachment?", "In the past 12 months when using a computer or laptop: have you installed a new device (hardware)?", "In the past 12 months when using a computer or laptop: have you searched for, downloaded, installed and configured software?","In the past 12 months when using a computer or laptop: have you created a digital presentation such as powerpoint, prezi or slideshare?", "In the past 12 months when using a computer or laptop: have you transferred folders between devices through the use of other devices such as USB-sticks?")
responses <- c("1 = yes, 2 = no", rep("1 = yes, 2 = no, 99 = does not know or no answer", 7))
table <- data.frame(cbind(code, form.section, question, responses))
names(table) <- names
kable(table, caption = "Code Book Table 4")
```

#### Digital Literacy of Members of Households with Learners

First, a subset of the data frame needs to be taken with respondents who (i) used a computer and (ii) are over the age of 14. In other cases, the digital literacy questions listed in the codebook were **not** asked. 

```{r}
df1 <- df %>%
  filter(s7p6 == 1 & s1p3 > 14)
```

Next, the table is pivoted (long) and appropriate names are given to the performed operations listed in the codebook. The responses for each performed operation are stored in the *Value* variable. The assumption is made that if respondents answered "I don't know" to a question, then they did *not* perform that action. This is a reasonable assumption, as they plausibly would understand the question if they had performed the operation in the past. 

```{r}
df2 <- df1 %>%
  select(id_per, area, "Copied or Moved Document/Folder" = s7p71, "Used Copy/Paste Function"= s7p72, 
         "Sent Emails with Attachment" = s7p73, "Installed New Devices" = s7p75, 
         "Downloaded and installed software" = s7p76,"Created a Digital Presentation" = s7p77, 
         "Used a USB-Stick" = s7p78) %>%
  pivot_longer(-c(id_per, area), names_to = "Action", values_to = "Value") %>%
  mutate(Value = factor(Value, levels = c(1,2,99), labels = c("yes", "no", "no"))) %>%
  mutate(Action = factor(Action))
```

The subsequent code calculates the proportion of respondents which performed different operations for rural and urban areas. The mean of the urban and rural proportion of respondents who performed a certain operation is used to order the operations.
```{r}
df3 <- df2 %>%
  group_by(area, Action, Value) %>%
  count() %>%
  ungroup() %>%
  group_by(area, Action) %>%
  mutate(prop = n/sum(n)*100) %>%
  filter(Value == "yes") %>%
  ungroup() %>%
  group_by(Action) %>%
  mutate(order = mean(prop))
```

The results are listed in *table 2*. 

```{r echo = FALSE, results = 'asis'}
table <- df3 %>%
        mutate(prop = paste0(round(prop), "%")) %>%
        select(-c(Value,n)) %>%
        pivot_wider(names_from = c(area), values_from = c(prop)) %>%
        arrange(-order) %>%
        select(-order)
kable(table, caption = "Table 2: Digital Actions Performed per Area")
```

Finally, a flipped bar chart is created which compares the proportions in terms of rural and urban respondents. 

```{r}
p <- ggplot(data = df3, aes(x = reorder(Action, order), y = prop, fill = area)) + theme_bw() 
p1 <- p + geom_bar(position = "dodge", stat = "identity") + coord_flip() 
p2 <- p1 + labs(y = "Proportion of respondents who\nperformed the action (in %)", x = "", 
                title = "Figure 3: Digital Actions Performed by Ecuadorian\nComputer Users in the last 12 Months", 
                subtitle = "Based on data from 7,808 respondents who: (1) lived in a household\nwith children and young people of school-going age; (2) had access\nto a computer in the past 12 months and (3) were 15 years or older.", 
                caption = "Source: Data from INEC Multi-Purpose Household Survey 2019\nData Summary by author (Max Aantjes)", fill = "")
p3 <- p2 +  scale_fill_manual(name = "", values = c("urban" = "#907D73", "rural" = "#C1C37F"))
p3
```

# References  

Andrade D., Bucheli V., Nabernegg M., & Garcés Christian, (2020). *Documento Metodológico de Diseño Muestral de la Encuesta Multipropósito*. INEC. Available from: https://www.ecuadorencifras.gob.ec/documentos/web-inec/Multiproposito/2019/201912_Metodologia%20Disenio%20Muestral_Multiproposito.pdf [accessed 23 August 2020]. 

INEC, (2019). *Encuesta Nacional Multipróposito de Hogares-Diciembre-2019*. Available from: https://www.ecuadorencifras.gob.ec/documentos/web-inec/Multiproposito/2019/201912_Formulario_Multiproposito.pdf [accessed 23 August 2020].  